\section{Geometric Brownian Motion}

Let $(X_t: t \ge 0)$ be a Brownian motion with constant drift on $\mbb{R}$ with generator

\begin{equation*}
    (\mathcal{L}f)(x) = \mu f'(x) + \frac{1}{2} \sigma^2 f''(x), \, \mu \in \mbb{R},\, \sigma > 0,
\end{equation*}

and initial condition $X_0 = 0$. \textbf{Geometric Brownian motion} is defined as

\begin{equation*}
    (Y_t: t \ge 0) \quad \text{with} \quad Y_t = e^{X_t}.
\end{equation*}

\begin{enumerate}
    \item[(a)] Show that $(Y_t: t \ge 0)$ is a diffusion process on $[0, \infty)$ and compute its generator. Write down the associated SDE and Fokker-Planck equation.

        \textit{ Sol. }

    \item[(b)] Use the evolution equation of expectation values of test functions $f: \mbb{R} \to \mbb{R}$
    
        \begin{equation*}
            \frac{\dif}{\dif t} \mbb{E}[f(Y_t)] = \mbb{E}[\mathcal{L}f(Y_t)],
        \end{equation*}

        to derive ODEs for the meane $m(t) := \mathbb{E}[Y_t]$ and the second moment $m_2(t) := \mbb{E}[Y_t^2]$. (No need to solve the ODEs.)

        \textit{ Sol. }

    \item[(c)] Under which conditions on $\mu$ and $\sigma^2$ is $(Y_t: t \ge 0)$ a martingale? 
    
    What is the asymptotic behaviour of the variance $v(t) = m_2(t) - m(t)^2$ in that case?
    
        \textit{ Sol. }

    \item[(d)] Show that $\delta_0$ is the unique stationary distribution of the process on the state space $[0, \infty)$. Under which conditions on $\mu$ and $\sigma^2$ does the process with $Y_0 = 1$ converge to the stationary distribution?
    
    Under which conditions on $\mu$ and $\sigma^2$ is the process ergodic? Justify your answer.

        \textit{ Sol. }

    \item[(e)] For $\sigma^2 = 1$ choose $\mu = -1/2$ and two other values $\mu < -1/2$ and $\mu > 1/2$. Simulate and plot a sample path of the process with $Y_0 = 1$ up to time $t = 10$, by numerically integrating the corresponding SDE with time steps $\Delta t = 0.1$ and $0.01$.
    
        \textit{ Sol. }

\end{enumerate}